Estimation of the perceived time of presence of sources in urban acoustic environments using deep learning techniques

### 2019-06-18 paper/paperV1.tex
Added links

\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{fancyhdr}
\usepackage[hidelinks]{hyperref}
\usepackage{graphicx}
\usepackage{epstopdf}
\usepackage{pdflscape}
\usepackage{multirow}
\usepackage{subcaption}
\usepackage{bbm}
\usepackage{color}
\DeclareMathOperator*{\argmax}{arg\,max}
\newcommand{\ml}[1]{\textcolor{red}{ML : #1}}
\begin{document}
\author{F\'elix Gontier $^1$, Catherine Lavandier $^2$, Pierre Aumond $^{2, 3}$\\Mathieu Lagrange $^1$, Jean-Francois Petiot $^1$}
\date{
$^1$ LS2N, UMR 6004 CNRS, Ecole Centrale de Nantes, F-44321, France\\
$^2$ ETIS, UMR 8051, University of Paris Seine, University of Cergy-Pontoise, ENSEA, CNRS, F-95000, France\\
$^3$ IFSTTAR, CEREMA, UMRAE, F-44344, Bouguenais, France
}
\title{Estimation of the perceived time of presence of sources in urban acoustic environments using deep learning techniques}
\maketitle

\begin{abstract}
The impact of urban sound on human beings has often been studied from a negative point of view (sound pollution), but in the two last decades with the soundscape approach, the positive impact has been revealed (resourcing spaces). The literature shows that the recognition of sources plays a great role in the way humans are affected by sound environments. There is thus a need for characterizing urban acoustic environments not only with sound pressure measurements but also with source-specific attributes such as their perceived time of presence. Combined, they allow us to predict important affective dimensions of the soundscape such as its pleasantness. However, computationally predicting source-specific attributes from recordings of sensors is notably harder than computing sound pressure levels.

This paper demonstrates, on a controlled dataset, that machine learning techniques based on state of the art neural architectures can predict the perceived time of presence of several sound sources with excellent performance, at a degree of precision that is sufficient to predict perceived pleasantness. To do so, a corpus of simulated sound scenes is first designed. Perceptual attributes corresponding to those stimuli are gathered through a listening experiment. The availability of the contributions of the individual sound sources within the simulated sound scenes then allows us to train and validate the proposed estimation procedure.

% to yield similar perceptual responses compared to recorded sound scenes.
PACS numbers: 43.66.Lj (Perceptual effects of sound), 43.60.-c (Acoustic signal processing)

\end{abstract}

@@

More precisely, the contributions of this paper are three-fold:
\begin{enumerate}
  \item Provide a corpus of simulated scenes for which relevant acoustic properties are well controlled\footnote{Corpus available at \url{} \ml{mettre sur archive}}.
  \item Propose numerical means~\footnote{Open source code is available at \url{}. \ml{mettre url}} for predicting the perceived time of presence of sound sources from raw acoustic data using deep learning approaches.
  \item Demonstrate through a comparative study with state of the art approaches that the proposed method allows the prediction of pleasantness to a sufficient degree.
\end{enumerate}

@@

Adjustments are made on the variance of the considered properties to better cover possible scenarios. Since voice events consist mostly of read English recordings, thus their EBR is reduced to improve realism. An additional \textit{square} ambiance is added with properties derived empirically from available data. Diverse new scenarios are then generated by sampling the background and event properties distributions, resulting in 40 scenes per ambiance for a total of 200 scenes. For each scene the perceptual time of presence for the traffic, voice and bird sources is estimated using the indicators proposed in~\cite{gontier2018}. Ideally, sound scenes should cover the resulting 3-dimensional space in a homogeneous way. To do so, the 75 scenes that maximize the minimum pairwise distance in the 3-dimensional space are selected, as shown in Figure~\ref{fig:tvb_pres}. The playback sound level is also conditioned to the ambiance according to typical sound levels in urban context, and ranges from 46.6~dB SPL to 77.1~dB SPL.

% Though, the time of presence is not a parameter of the simulation process and thus the distribution of the scene population in this space is not controlled. As an alternative,

@@

The perception of source activity can be linked to the notions of dominance, time of presence, or emergence. The perceptual time of presence, that is the percentage of time for which the corresponding source is heard, is selected as the activity descriptor for this study. The influence of traffic on pleasantness is further separated into that of background and salient events. Thus, 4 additional questions are presented to the participants and evaluated on the same 11 point scales:
\begin{itemize}
\item Sound level of passing vehicles: \textit{Very low - Very high} (\textit{Tr\`es faible - Tr\`es fort}) - $L_{T, p}$,
\item Time of presence of Traffic, Voices, Birds: \textit{Never - Continuously} (\textit{Jamais - Continuellement}), resp. $T_{T, p}$, $T_{V, p}$ and $T_{B, p}$
\end{itemize}
where $p$ denotes a perceptual evaluation.

@@

The scenes are played at a given sound level as discussed in Section~\ref{sec:data_corp}, through the same computer and sound card configurations. Beyerdynamics DT-990 Pro headphones are used for all participants. These headphones are calibrated in a semi-anechoic chamber where a relation between the playback sound level and the sound card output voltage is obtained for pink noise. From this information a scaling factor is applied to the sound scenes to ensure that they are heard at the desired sound level by every listener.

A total of 23 students aged from 22 to 23 years including 16 males and 7 females at Ecole Centrale de Nantes completed the test. All gave consent prior to the experiment and reported normal hearing. With 23 subjects, the incomplete block design is not perfectly balanced.

@@

%Participants evaluate each scene on criteria using 11-point Likert scales. Many statistical tools assume that the studied data follows continuous, single mode distributions. It is thus important to first verify that no distinct groups of participants can be obtained from perceptual assessments. A hierarchical clustering analysis of the judgements of the participants is performed for the 25 scenes in the commonly evaluated sub-corpus. Starting with the assessments of each participant as separate clusters, new clusters are formed iteratively where the average euclidean distance between all pairs within the two merged clusters at each step is minimal. and shown in Figure~\ref{fig:hclusters}. No clear separation into groups is found, as a result the assessment distributions are assumed to be single-mode and their means can be used for further analysis.\\
%
%\begin{figure}[!h]
%    \centering
%    \includegraphics[width=0.7\textwidth]{figures/subj_gr.eps}
%    \caption{Dendrogram of the hierarchical clustering of participants on the first 25 scenes of the test. No distinct groups are visible.}\label{fig:hclusters}
%\end{figure}

The perceptual responses of the 6 recorded and corresponding replicated scenes are first studied. Table~\ref{tab:ogrep} shows the mean differences between assessments for pairs of scenes with equivalent scenarios. Wilcoxon signed-rank tests~\cite{wilcoxon1945} are implemented for each scene and perceptual scale to outline significant differences between assessment distributions, which are shown in bold. As the data is discretely distributed, zero differences between paired samples are included using Pratt's modification of the test~\cite{pratt1959}. On the general scales (P, L, OL, I and C), all mean differences are lower than 2 points on the 11-point Likert scale. Though, significant differences are outlined that can be linked to corresponding discrepancies in source-specific parameters. The highest difference (-5.22) is found for the assessment of the time of presence of traffic in the location P4. For this location, the background traffic in the recorded scene varies along time, it is louder in the first half of the scene than in the second half. Replicating this scene using simScene imposes a constant sound level for background sources. Thus, the background traffic is louder in the replicated scene than it is in the recording for about half of its duration. To a lesser extent the same issue explains the large difference (-4) in the assessed time of presence of voices in the location P3. Discrepancies on source-specific scales can also be interpreted by the choice of isolated samples, which is semi-random and based on a high-level source taxonomy. For example, no difference is made during annotation between child or adult speech, or depending on its expressiveness. Overall, no consistent difference between the perception of recorded and replicated scenes emerges for the studied points. Thus, the proposed hypothesis that recorded and replicated scenes generate equivalent perceptual responses cannot be rejected.

\begin{table*}[t]
\centering
\caption{Mean differences of perceptual assessments between original and replicated sound scenes. Significantly different distributions as per a Wilcoxon signed-rank test are shown in bold (n=23, p<0.05)}
\label{tab:ogrep}
%\resizebox{\columnwidth}{!}{
\begin{tabular}{ c | c c c c c c c c c }
\hline
	 & P & L & OL & I & C & $L_{T, p}$ & $T_{T, p}$ & $T_{V, p}$ & $T_{B, p}$ \\ \hline
	P1 & 0.43 & \textbf{-1.65} & \textbf{-1.04} & 0.43 & 0.13 & \textbf{-0.91} & 0.39 & \textbf{-2.09} & 0.61 \\
	P3 & 0.26 & -0.43 & 0.30 & -1 & 0.30 & 0.35 & 1.04 & \textbf{-4} & 0.22 \\
	P4 & 0.91 & 0 & \textbf{-1.83} & 0.48 & \textbf{1.30} & \textbf{-0.96} & \textbf{-5.22} & \textbf{1.43} & 0.04 \\
	P8 & 0.26 & \textbf{-1.65} & -0.87 & -0.96 & 0.65 & \textbf{-2.04} & \textbf{-0.91} & 0.09 & \textbf{-1.43} \\
	P15 & \textbf{-1.35} & 0.52 & 0.52 & \textbf{-1.17} & 0.09 & 0.61 & 0.13 & \textbf{1.96} & \textbf{-2.74} \\
	P18 & \textbf{1.13} & -0.30 & \textbf{-1.17} & -0.43 & \textbf{1.39} & -1.04 & \textbf{-1.83} & \textbf{0.83} & \textbf{1.30} \\ \hline
\end{tabular}
%}
\end{table*}

To validate the use of simulated sound scenes with new scenarios as well as reduced source complexity in the development of prediction models, the perceptual space generated by the experiment's five general scales (pleasantness, liveliness, overall loudness, interest and calmness) is studied. It is obtained by performing a principal components analysis (PCA) on the corresponding perceptual responses averaged along participants. No standardization is applied to the data. Figure~\ref{fig:pspace_rec} and Figure~\ref{fig:pspace_sim} compare the results for scenes based on recordings (n=25) and new scenarios (n=75) respectively. The resulting spaces are similar, with only overall loudness and pleasantness axes slightly rotated between the two subcorpora. For both sets the variance explained by the first two components is similar, resp. 79.4\% - 18.1\% and 79.6\% - 15.2\%. Furthermore, these representations are comparable to those found in previous work on perceptual dimensions~\cite{axelsson2010, cain2013}. Thus, the use of simulated scenes based on both real and new scenarios does not result in major differences in the relations between perceptual quantities. Additionally, the assessments averaged on all subjects are projected onto the PCA space for each scene and represented in Figure~\ref{fig:pspace_sim} as dots for simulated scenes corpus and crosses for recorded and replicated scenes. The space covered by scenes based on new scenarios covers that of the studied real-life environments, further demonstrating the effectiveness of the scene generation procedure in terms of diversity of scenarios.

Discrepancies between projections of original and replicated scenes are highlighted in Figure~\ref{fig:pspace_rec} using arrows. The standard deviation of assessments is represented using ellipses for one location (P1). The projections of assessment distributions consistently overlap for all pairs of recorded and replicated scenes\footnote{Data available here \url{} \ml{mettre url}}. This illustrates the results in Table~\ref{tab:ogrep} where perceptual assessments were not found to differ significantly overall.

\ml{On vise 12 pages. Tu dois raccourcir les legendes des figures.}

@@

\begin{align}
TFSD_{f, t} = \frac{\lvert\frac{d^2L(f, t)}{dfdt}\rvert}{\sum_{f1=31.5Hz}^{f1=16kHz}\lvert\frac{d^2L(f, t)}{dfdt}\rvert}
\end{align}

where $L(f, t)$ is the third-octave spectrum of the signal. To provide a baseline for pleasantness prediction, the $TFSD$ is computed for the 4kHz band and 125~ms measurements ($TFSD_{4kHz(1/8s)}$) as an estimate the contribution of birds. The 500Hz band with 1~s measurements ($TFSD_{500Hz(1s)}$) is chosen as an estimate the contribution of voices.\\

The simulation process of the scenes outputs ground truth source contributions as separate channels. Additional indicators are computed on the separated audio signals for the traffic, voices and birds sources:

\begin{itemize}
\item $L_{eq, s}$: Equivalent sound level for source $s$,
\item $\Delta L_{s}$: Source emergence, taken as the difference between the equivalent sound level of source $s$ and that of all other sources combined.
\end{itemize}

Next, the $\hat T_s(\alpha, \beta)$ time of presence approximation proposed in~\cite{gontier2018} is computed. $\hat T_s(\alpha, \beta)$ is based on a binary source emergence model computed on the third-octave band emergence $\Delta L_{s}(t, f) = L_s(t, f) - L_{\bar{s}}(t, f)$. It is parametrized by $\alpha$ and $\beta$ thresholds:

@@

where $r$ is the Pearson's correlation coefficient, $s$ denotes the sound source, $N_s$ is the number of sources in the taxonomy and equals 3 in this study, $t$ is the time frame and $T_{s, p}$ corresponds to the perceptual time of presence assessments averaged per scene. This indicator evaluates the presence or absence of a given source in a small time frame. First, frequency bands for which the source is emergent by more than the $\alpha$ threshold value are isolated. Then, the source is considered present if the emergence for these bands is on average greater than the $\beta$ threshold. A time of presence estimation is obtained for each source by averaging along the $N_t$ time frames composing the scene. The optimal threshold values are optimized once in this study on the whole perceptual experiment corpus, and found as $\alpha_{opt} = -14dB$ and $\beta_{opt} = -7dB$. This means that to be considered heard in a given time frame, the source has to be emergent by more than $-7dB$ on average on third-octave bands where its level is at least $14dB$ below other sources.

@@

A deep learning model is implemented and trained for the source presence prediction task, with the architecture shown in Figure~\ref{fig:deep_arch}. The model includes 4 blocks of convolutional layers followed by leaky rectified linear unit (LeakyReLU) activations of expression $y = max(0.1x, x)$. The convolutional layers have respectively 128, 64, 32 and 32 output maps, \textit{i.e.} channels, and a common kernel size of 5x5. The output of the last block is flattened then goes through a fully connected layer with output size 3. A final sigmoid activation is used in order to obtain outputs in the 0-1 range. The resulting 3 values correspond to the presence of traffic, voices and birds in the 1~s frame respectively. During training these values are directly compared to presence labels given by $\hat T_s(\alpha_{opt}, \beta_{opt})$ using a binary cross-entropy cost function:

\begin{align}
BCE(y, \hat y) = -\sum_s y_s log\left(\hat y_s\right) + (1-y_s) log\left(1-\hat y_s\right)
\end{align}

where $s$ is the source, $y_s$ and $\hat y_s$ are the target and predicted presence for source $s$ in the 0-1 range. This loss function is minimized using the Adam algorithm~\cite{kingma2015} on batches of 1~s examples. During evaluation, a threshold of 0.5 is independently applied to the 3 outputs to obtain a binary presence value for each source: each source is considered absent when the model outputs a value lower than 0.5 and present when it outputs 0.5 or higher. The time of presence estimation is then obtained by averaging presence labels of all 1~s time frames corresponding to the same scene.

@@

As a baseline, multilinear models of pleasantness are built as a function of the overall loudness (OL) and source-specific parameters ($L_{T, p}$, $T_{T, p}$, $T_{V, p}$, $T_{B, p}$) with assessments averaged for each scene (n=100). All 31 combinations of the 5 predictors are considered, but to ensure that no multi-collinearity is present between predictors a variance inflation factor (VIF) test is performed prior to each regression. Only combinations for which all predictors verify $VIF<5$ are considered valid. The best resulting models in terms of $R^2_{adj}$ with statistically significant coeffeicient estimates ($p<0.05$) are:

\begin{align}
\hat P_{1, p} &= 8.56 - 0.63~OL - 0.20~L_{T, p} + 0.11~T_{V, p} + 0.14~T_{B, p} \label{eq:pp1}\\
\hat P_{2, p} &= 8.99 - 0.67~OL - 0.15~T_{T, p} + 0.08~T_{V, p} + 0.12~T_{B, p} \label{eq:pp2}
\end{align}

@@

Table~\ref{tab:percm} summarizes performance metrics computed for the two models. $\hat P_{1, p}$ and $\hat P_{2, p}$ display similar estimation errors, adjusted $R^2$ and correlation to assessed pleasantness. The performance of obtained perceptual models is about 0.6 RMSE, which is below the mean standard deviation of pleasantness assessments for this experiment (1.77).

@@

Table~\ref{tab:physc} shows the Pearson's correlation coefficients between computed physical indicators and perceptual assessments. First, all global sound level indicators correlate well (r>0.9) with the perceived overall loudness. Regarding source-specific perceptual parameters, the $L_{eq, s}$ correlates consistently well with the $T_{s, p}$ of corresponding source (r=0.71). Conversely, the emergence $\Delta L_s$ fails to represent the perceived bird activity, and correlations are weak for other sources. The proposed estimates of the time of presence $\hat T_s(\alpha_{opt}, \beta_{opt})$ show strong correlations to their perceptual counterparts (r>0.8), though this is expected as $\alpha_{opt}$ and $\beta_{opt}$ are optimized to this aim. They also display good source discrimination properties, as no significant correlation is found between voices and birds, and perceptual assessments of traffic were correlated with those of other sources in Table~\ref{tab:percc}. They are also better predictors than the $TFSD_{500Hz, 1s}$ and $TFSD_{4kHz, 1/8s}$ indicators for voice and bird activity in this corpus. Interestingly, the perceived sound level of passing vehicles $L_{T, p}$ is best represented by the $L_{10}$. In perceptual models including the sound level of traffic events, the global $L_{10}$ may thus be a sufficient predictor without a source separation and level estimation algorithm.

@@

\begin{align}
\hat P_{1, \varphi} = 16.74 - 0.18~L_{50} + 1.01~\hat T_B(\alpha_{opt}, \beta_{opt}) \label{eq:pphi}
\end{align}

@@

\begin{align}
\hat P_{2, \varphi} &= 18.67 - 0.20~L_{50} - 0.02~(L_{10}-L_{90})\\
\begin{split}
\hat P_{3, \varphi} &= 30.18 - 0.16~L_{50, 1kHz} + 8.92~TFSD_{500Hz, 1s} \\
& + 2.99~TFSD_{4kHz, 1/8s}
\end{split}
\end{align}

@@

Pleasantness prediction for simulated scenes using the detection model is as effective as the best baseline model from Section~\ref{sec:results_phys}, and outperforms previous models relying on physical indicators directly computed on the audio signal. %Though, perceptual baseline models presented in Section~\ref{sec:results_perc} still display better overall performance metrics for all considered sub-corpora.

@@

\begin{align}
\begin{split}
\hat P_{3, p} {}&= 7.11 - 0.38~OL - 0.15~T_{T, p} + 0.20~T_{V, p} \\
& + 0.15~T_{B, p}
\end{split}\\
\begin{split}
\hat P_{4, p} {}&= 8.70 - 0.47~OL - 0.21~T_{T, p} + 0.12~T_{V, p} \\
& + 0.09~T_{B, p}
\end{split}
\end{align}

>>>>

\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{fancyhdr}
\usepackage[hidelinks]{hyperref}
\usepackage{graphicx}
\usepackage{epstopdf}
\usepackage{pdflscape}
\usepackage{multirow}
\usepackage{subcaption}
\usepackage{bbm}
\usepackage{color}
\DeclareMathOperator*{\argmax}{arg\,max}
\newcommand{\ml}[1]{\textcolor{red}{ML : #1}}
\begin{document}
\author{F\'elix Gontier $^1$, Catherine Lavandier $^2$, Pierre Aumond $^{2, 3}$\\Mathieu Lagrange $^1$, Jean-Francois Petiot $^1$}
\date{
$^1$ LS2N, UMR CNRS 6004, Ecole Centrale de Nantes, F-44321, France\\
$^2$ ETIS, UMR CNRS 8051, University of Paris Seine, University of Cergy-Pontoise, ENSEA, CNRS, F-95000, France\\
$^3$ IFSTTAR, CEREMA, UMRAE, F-44344, Bouguenais, France
}
\title{Estimation of the perceived time of presence of sources in urban acoustic environments using deep learning techniques}
\maketitle

\begin{abstract}
The impact of urban sound on human beings has often been studied from a negative point of view (sound pollution), but in the two last decades with the soundscape approach, the positive impact has been revealed (resourcing spaces). The literature shows that the recognition of sources plays a great role in the way humans are affected by sound environments. There is thus a need for characterizing urban acoustic environments not only with sound pressure measurements but also with source-specific attributes such as their perceived time of presence. Combined, they allow the prediction of important affective dimensions of the soundscape such as its pleasantness. However, computationally predicting source-specific attributes from recordings of sensors is notably harder than computing sound pressure levels.

This paper demonstrates, on a controlled dataset, that machine learning techniques based on state of the art neural architectures can predict the perceived time of presence of several sound sources with excellent performance, at a degree of precision that is sufficient to predict perceived pleasantness. To do so, a corpus of simulated sound scenes is first designed. Perceptual attributes corresponding to those stimuli are gathered through a listening experiment. The availability of the contributions of the individual sound sources within the simulated sound scenes then allows the training and validate the proposed estimation procedure.

PACS numbers: 43.66.Lj (Perceptual effects of sound), 43.60.-c (Acoustic signal processing)
\end{abstract}

@@

More precisely, the contributions of this paper are three-fold:
\begin{enumerate}
  \item Provide a corpus of simulated scenes for which relevant acoustic properties are well controlled\footnote{Corpus available at \url{https://zenodo.org/record/3248734\#.XQjC4v7gqUk}}.
  \item Propose numerical means~\footnote{Open source code is available at \url{https://github.com/felixgontier/soundSourcePresenceEstimation}.} for predicting the perceived time of presence of sound sources from raw acoustic data using deep learning approaches.
  \item Demonstrate through a comparative study with state of the art approaches that the proposed method allows the prediction of pleasantness to a sufficient degree of accuracy.
\end{enumerate}

@@

Adjustments are made on the variance of the considered properties to better cover possible scenarios. Since voice events consist mostly of read English recordings, their EBR is reduced to improve realism. An additional \textit{square} ambiance is added with properties derived empirically from available data. Diverse new scenarios are then generated by sampling the background and event properties distributions, resulting in 40 scenes per ambiance for a total of 200 scenes. For each scene the perceptual time of presence for the traffic, voice and bird sources is estimated using the indicators proposed in~\cite{gontier2018}. Ideally, sound scenes should cover the resulting 3-dimensional space in a homogeneous way. To do so, the 75 scenes that maximize the minimum pairwise distance in the 3-dimensional space are selected, as shown in Figure~\ref{fig:tvb_pres}. The playback sound level is also conditioned to the ambiance according to typical sound levels in urban context, and ranges from 46.6~dB SPL to 77.1~dB SPL.

@@

The perception of source activity can be linked to the notions of dominance, time of presence, or emergence. The perceptual time of presence, that is the percentage of time for which the corresponding source is heard, is selected as the activity descriptor for this study. The influence of traffic on pleasantness is further separated into that of background and salient events. Thus, 4 additional questions are presented to the participants and evaluated on the same 11 point scales:
\begin{itemize}
\item Sound level of passing vehicles: \textit{Very low - Very high} (\textit{Tr\`es faible - Tr\`es fort}) - $L_{T, p}$,
\item Time of presence of Traffic, Voices, Birds: \textit{Never - Continuously} (\textit{Jamais - Continuellement}) - resp. $T_{T, p}$, $T_{V, p}$ and $T_{B, p}$
\end{itemize}
where $p$ denotes a perceptual evaluation.

@@

The scenes are played at a given sound level as discussed in Section~\ref{sec:data_corp}, through the same computer and sound card configurations. Beyerdynamics DT-990 Pro headphones are used by all participants. These headphones are calibrated in a semi-anechoic chamber where a relation between the playback sound level and the sound card output voltage is obtained for pink noise. From this information a scaling factor is applied to the sound scenes to ensure that they are heard at the desired sound level by every listener.

A total of 23 students aged from 22 to 23 years including 16 males and 7 females at Ecole Centrale de Nantes completed the test (note that with 23 subjects, the incomplete block design is not perfectly balanced). All gave consent prior to the experiment and reported normal hearing.

@@

The perceptual responses of the 6 recorded and corresponding replicated scenes are first studied. Table~\ref{tab:ogrep} shows the mean differences between assessments for pairs of scenes with equivalent scenarios. Wilcoxon signed-rank tests~\cite{wilcoxon1945} are implemented for each scene and perceptual scale to outline significant differences between assessment distributions, which are shown in bold. As the data is discretely distributed, zero differences between paired samples are included using Pratt's modification of the test~\cite{pratt1959}. On the first five scales, all mean differences are lower than 2 points on the 11-point Likert scale. Though, significant differences are outlined that can be linked to corresponding discrepancies in source-specific parameters. The highest difference (-5.22) is found for the assessment of the time of presence of traffic in the location P4. For this location, the background traffic in the recorded scene varies along time, it is louder in the first half of the scene than in the second half. Replicating this scene using simScene imposes a constant sound level for background sources. Thus, the background traffic is louder in the replicated scene than it is in the recording for about half of its duration. To a lesser extent the same issue explains the large difference (-4) in the assessed time of presence of voices in the location P3. Discrepancies on source-specific scales can also be interpreted by the choice of isolated samples, which is semi-random and based on a high-level source taxonomy. For example, no difference is made during annotation between child or adult speech, or depending on its expressiveness. Though, overall no consistent difference between the perception of recorded and replicated scenes emerges for the studied points.

\begin{table*}[t]
\centering
\caption{Mean differences of perceptual assessments between original and replicated sound scenes. Significant differences as per a Wilcoxon signed-rank test are shown in bold (n=23, p<0.05)}
\label{tab:ogrep}
%\resizebox{\columnwidth}{!}{
\begin{tabular}{ c | c c c c c c c c c }
\hline
	 & P & L & OL & I & C & $L_{T, p}$ & $T_{T, p}$ & $T_{V, p}$ & $T_{B, p}$ \\ \hline
	P1 & 0.43 & \textbf{-1.65} & \textbf{-1.04} & 0.43 & 0.13 & \textbf{-0.91} & 0.39 & \textbf{-2.09} & 0.61 \\
	P3 & 0.26 & -0.43 & 0.30 & -1 & 0.30 & 0.35 & 1.04 & \textbf{-4} & 0.22 \\
	P4 & 0.91 & 0 & \textbf{-1.83} & 0.48 & \textbf{1.30} & \textbf{-0.96} & \textbf{-5.22} & \textbf{1.43} & 0.04 \\
	P8 & 0.26 & \textbf{-1.65} & -0.87 & -0.96 & 0.65 & \textbf{-2.04} & \textbf{-0.91} & 0.09 & \textbf{-1.43} \\
	P15 & \textbf{-1.35} & 0.52 & 0.52 & \textbf{-1.17} & 0.09 & 0.61 & 0.13 & \textbf{1.96} & \textbf{-2.74} \\
	P18 & \textbf{1.13} & -0.30 & \textbf{-1.17} & -0.43 & \textbf{1.39} & -1.04 & \textbf{-1.83} & \textbf{0.83} & \textbf{1.30} \\ \hline
\end{tabular}
%}
\end{table*}

Next, the perceptual space generated by the experiment's five general scales (P, L, OL, I and C) is studied to validate the use of simulated sound scenes with new scenarios as well as reduced source complexity. It is obtained by performing a principal components analysis (PCA) on the corresponding perceptual responses averaged along participants. No standardization is applied to the data. Figure~\ref{fig:pspace_rec} and Figure~\ref{fig:pspace_sim} compare the results for scenes based on recordings (n=25) and new scenarios (n=75) respectively. The resulting spaces are similar, with only overall loudness and pleasantness axes slightly rotated between the two subcorpora. For both sets the variance explained by the first two components is similar, resp. 79.4\% - 18.1\% and 79.6\% - 15.2\%. Furthermore, these representations are comparable to those found in previous work on perceptual dimensions~\cite{axelsson2010, cain2013}. Thus, the use of simulated scenes based on both real and new scenarios does not result in major differences in the relations between perceptual quantities. Additionally, the assessments averaged on all subjects are projected onto the PCA space for each scene and represented in Figure~\ref{fig:pspace_sim} as dots for simulated scenes corpus and crosses for recorded and replicated scenes. These projections show that the space covered by scenes based on new scenarios covers that of the studied real-life environments. This further demonstrates the effectiveness of the scene generation procedure in terms of diversity of scenarios.

Discrepancies between projections of original and replicated scenes are highlighted in Figure~\ref{fig:pspace_rec} using arrows. The standard deviation of assessments is represented using ellipses for one location (P1). The projections of assessment distributions consistently overlap for all pairs of recorded and replicated scenes\footnote{Data available here \url{http://felixgontier.github.io/soundSourcePresenceEstimation/index.html}}. This illustrates the results in Table~\ref{tab:ogrep} where perceptual assessments were not found to differ significantly overall.

@@

\begin{equation}
TFSD_{f, t} = \frac{\lvert\frac{d^2L(f, t)}{dfdt}\rvert}{\sum_{f1=31.5Hz}^{f1=16kHz}\lvert\frac{d^2L(f1, t)}{df1dt}\rvert}
\end{equation}
where $L(f, t)$ is the third-octave spectrum of the signal. To provide a baseline for pleasantness prediction, the $TFSD$ is computed for the 4kHz band and 125ms measurements ($TFSD_{4kHz(1/8s)}$), and for the 500Hz band with 1s measurements ($TFSD_{500Hz, 1s}$) as estimates of the contribution of birds and voices respectively.

The simulation process of the scenes outputs ground truth source contributions as separate channels. Additional indicators are computed on the separated audio signals for the traffic, voices and birds sources: The equivalent sound level $L_{eq, s}$ for source $s$ and the source emergence $\Delta L_{s}$, taken as the difference between the equivalent sound level of source $s$ and that of all other sources combined.

Next, the $\hat T_s(\alpha, \beta)$ time of presence approximation proposed in~\cite{gontier2018} is computed. $\hat T_s(\alpha, \beta)$ is based on a binary source emergence model computed on the third-octave band emergence spectrum $\Delta L_{s}(t, f)$. It is parametrized by $\alpha$ and $\beta$ thresholds:

@@

where $r$ is the Pearson's correlation coefficient, $s$ denotes the sound source, $N_s$ is the number of sources in the taxonomy and equals 3 in this study, $t$ is the time frame and $T_{s, p}$ corresponds to the perceptual time of presence assessments averaged per scene. This indicator evaluates the presence or absence of a given source in a small time frame. First, frequency bands for which the source is emergent by more than the $\alpha$ threshold value are isolated. Then, the source is considered present if the emergence for these bands is on average greater than the $\beta$ threshold. A time of presence estimation is obtained for each source by averaging along the $N_t$ time frames composing the scene. The optimal threshold values are optimized once in this study on the whole perceptual experiment corpus, and found as $\alpha_{opt} = -14dB$ and $\beta_{opt} = -7dB$.

@@

A deep learning model is implemented and trained for the source presence prediction task, with the architecture shown in Figure~\ref{fig:deep_arch}. The model includes 4 blocks of convolutional layers followed by leaky rectified linear unit (LeakyReLU) activations of expression $y = max(0.1x, x)$. The convolutional layers have respectively 128, 64, 32 and 32 output channels, and a common kernel size of 5x5. The output of the last block is flattened then goes through a fully connected layer with output size 3. A final sigmoid activation is used in order to obtain outputs in the 0-1 range, which correspond to the presence of traffic, voices and birds in the 1~s frame respectively. During training these values are directly compared to presence labels given by $\hat T_s(\alpha_{opt}, \beta_{opt})$ using a binary cross-entropy cost function:

\begin{equation}
BCE(y, \hat y) = -\sum_s y_s log\left(\hat y_s\right) + (1-y_s) log\left(1-\hat y_s\right)
\end{equation}
where $s$ is the source, $y_s$ and $\hat y_s$ are the target and predicted presence for source $s$ in the 0-1 range. This loss function is minimized using the Adam algorithm~\cite{kingma2015} on batches of 1~s examples. During evaluation, a threshold of 0.5 is independently applied to the 3 outputs to obtain a binary presence value for each source: each source is considered absent when the model outputs a value lower than 0.5 and present when it outputs 0.5 or higher. The time of presence estimation is then obtained by averaging presence labels of all 1~s time frames corresponding to the same scene.

@@

As a baseline, multilinear models of pleasantness are built as a function of the overall loudness (OL) and source-specific parameters ($L_{T, p}$, $T_{T, p}$, $T_{V, p}$, $T_{B, p}$) with assessments averaged for each scene (n=100). All 31 combinations of the 5 predictors are considered, but to ensure that no multi-collinearity is present between predictors a variance inflation factor (VIF) test is performed prior to each regression. Only combinations for which all predictors verify $VIF<5$ are considered valid. The best resulting models in terms of $R^2_{adj}$ with statistically significant coefficient estimates ($p<0.05$) are:

\begin{align}
\begin{split}
\hat P_{1, p} & = 8.56 - 0.63~OL - 0.20~L_{T, p} + 0.11~T_{V, p}\\
&\qquad + 0.14~T_{B, p} \label{eq:pp1}
\end{split}\\
\begin{split}
\hat P_{2, p} & = 8.99 - 0.67~OL - 0.15~T_{T, p} + 0.08~T_{V, p}\\
&\qquad + 0.12~T_{B, p} \label{eq:pp2}
\end{split}
\end{align}

@@

Table~\ref{tab:percm} summarizes performance metrics computed for the two models. $\hat P_{1, p}$ and $\hat P_{2, p}$ display similar estimation errors, adjusted $R^2$ and correlation to assessed pleasantness. The performance of obtained perceptual models is about 0.6 (RMSE), which is below the mean standard deviation of pleasantness assessments for this experiment (1.77).

@@

Table~\ref{tab:physc} shows the Pearson's correlation coefficients between computed physical indicators and perceptual assessments. First, all global sound level indicators correlate well (r>0.9) with the perceived overall loudness. Regarding source-specific perceptual parameters, the $L_{eq, s}$ correlates consistently well with the $T_{s, p}$ of corresponding source (r=0.71). Conversely, the emergence $\Delta L_s$ fails to represent the perceived bird activity, and correlations are weak for other sources. The proposed estimates of the time of presence $\hat T_s(\alpha_{opt}, \beta_{opt})$ show strong correlations to their perceptual counterparts (r>0.8), though this is expected as $\alpha_{opt}$ and $\beta_{opt}$ are optimized to this aim. They also display good source discrimination properties, as no significant correlation is found between voices and birds, and perceptual assessments of traffic were correlated with those of other sources in Table~\ref{tab:percc}. They are also better predictors than the $TFSD_{500Hz, 1s}$ and $TFSD_{4kHz, 1/8s}$ indicators for voice and bird activity in this corpus. Interestingly, the perceived sound level of passing vehicles $L_{T, p}$ is best represented by the $L_{10}$. In perceptual models including the sound level of traffic events, the global $L_{10}$ may thus be a relevant predictor without the need for a source separation and level estimation algorithm.

@@

\begin{equation}
\hat P_{1, \varphi} = 16.74 - 0.18~L_{50} + 1.01~\hat T_B(\alpha_{opt}, \beta_{opt}) \label{eq:pphi}
\end{equation}

@@

\begin{align}
\hat P_{2, \varphi} &= 18.67 - 0.20~L_{50} - 0.02~(L_{10}-L_{90})\\
\begin{split}
\hat P_{3, \varphi} &= 30.18 - 0.16~L_{50, 1kHz} + 8.92~TFSD_{500Hz, 1s} \\
&\qquad + 2.99~TFSD_{4kHz, 1/8s}
\end{split}
\end{align}

@@

Pleasantness prediction for simulated scenes using the detection model is as effective as the best baseline model from Section~\ref{sec:results_phys}, and outperforms previous models relying on physical indicators directly computed on the audio signal.

@@

\begin{align}
\begin{split}
\hat P_{3, p} {}&= 7.11 - 0.38~OL - 0.15~T_{T, p} + 0.20~T_{V, p} \\
&\qquad + 0.15~T_{B, p}
\end{split}\\
\begin{split}
\hat P_{4, p} {}&= 8.70 - 0.47~OL - 0.21~T_{T, p} + 0.12~T_{V, p} \\
&\qquad + 0.09~T_{B, p}
\end{split}
\end{align}
